---
title: "Rowley_HLM2_Final"
author: "Brock Rowley"
date: "6/8/2021"
output:
  html_document:
    css: table_style.css
bibliography: r-references.bib
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(papaja)
library(tidyverse)
library(knitr)
library(kableExtra)
library(lme4)
library(here)
library(rio)
library(equatiomatic)
library(performance)
library(readr)
library(brms)
library(insight)
library(broom.mixed)
library(gghalves)

year7 <- rio::import(here::here("data", "Year7_MLM.sav"))
view(year7)
### Explanation of variables:
# * School = School ID.
# * SID = Student ID.
# * RaceEthnicity = Categorical variable of student race.
# * OSS_1stRef = True/False (logical) student received Out of School Suspension
# (OSS) on first referral in school year for substance possession.
# * Alcohol.1, Drugs.1, Tobacco.1 = True/False (logical) student possessed named
# substance in 1st referral.
# * subtype = categorical variable of substance possessed on first referral.
# Mutually exclusive categories eliminate some co-ocurrence that happened.
# Probably won't use this variable, but there if needed.
# * Gender = Male (0), Female(1)
# * ODRCount = # of Office Discipline Refferal (ODRs) for substance in that school
# year for each student.
# * Grade = grade of student at time of referral.
# * IEP = True/False (logical) if student had Individualize Education Plan (IEP)
# at time of referral.
# * BLW_Sample = Flag for students in schools that had minimum of 5% students in
# each abbreviated race category (e.g., B=Black, L=Latinx) race categories
# referred for substances in year. Likely not needed for these student-level
# analyses.
# * Tier1Fidelity: school was implementing Positive Behavioral Intervention
# Supports (PBIS) with fidelity (1) or not (0) based on one of three measures that
# year.
```

```{r clean_group}
year7$PrimarySubstanceType.1 <- as.factor(year7$PrimarySubstanceType.1)

year7 <- year7 %>%
   mutate(subtype = ifelse(PrimarySubstanceType.1 == 13, "Tobacco", 
                     ifelse(PrimarySubstanceType.1 == 14, "Alcohol", "Drugs")), 
          Alcohol.1 = as.factor(year7$Alcohol.1),
          Drugs.1 = as.factor(year7$Drugs.1),
          Tobacco.1 = as.factor(year7$Tobacco.1),
          RaceEthnicity = as.factor(year7$RaceEthnicity),
          School = as.factor(year7$School),
          District = as.factor(year7$District),
          RaceEthnicity = case_when(RaceEthnicity == 1 ~ "Native",
                                    RaceEthnicity == 2 ~ "Asian",
                                    RaceEthnicity == 3 ~ "Latinx",
                                    RaceEthnicity == 4 ~ "Black",
                                    RaceEthnicity == 5 ~ "White",
                                    RaceEthnicity == 8 ~ "Pacific_Islander",
                                    RaceEthnicity == 44 ~ "Multi"),
          RaceEthnicity = as.factor(year7$RaceEthnicity))

grouped <- year7 %>%
  group_by(RaceEthnicity) %>%
  count()
view(grouped)
```

# Introduction:
  Student substance misuse is an ongoing problem in schools across the United
States, mostly outside the training and expertise of most educators.
Nonetheless, school staff are presented with situations in which students bring
substances to campus or arrive under the influence. Despite the evidence that
zero-tolerance policies do not improve school safety or student behavior
(@americanpsychologist), the likely default response in
contemporary schools to substance-related behavior occurs through their
discipline systems. Effective support for students is most promising if
delivered through a preventative and proactive system as outlined with the
Interconnected Systems Framework (ISF) (@emoreview). The ISF advocates
integration and coordination with mental health service providers to meet a
broader range of behavioral health needs more effectively. Although the national
mindset is moving toward a more inclusive approach to behavioral health, this
study seeks to determine to what extent school practices have shifted with it.
The objective of this study is to leverage a national Office Discipline Referral
(ODR) data set to examine the administrative decisions for substance possession
referrals. Specifically, this study seeks to examine how use of Out-of-School
Suspension (OSS) differs by substance possessed and student race/ethnicity.

# Method:
  ODR data from the SWIS database were extracted following a formal, written data
request submitted by this author. This data set was obtained as a referral-level
data set and required conversion to a student-level data set by running cases to
variables command in SPSS. For all analyses of administrative actions taken in
response to referrals, only a student’s first referral for substance possession
was used, meaning any administrative actions in subsequent referrals were not
included. Analysis was done using only the first referrals for substance
possession for each student in the data set to eliminate concerns about possible
confounding of analyses related to decision-making with subsequent referrals.
## Data Source:
  The data sample for the project comes from the web-based School-Wide
Information System (SWIS) for collecting and analyzing school discipline data in
the form of office discipline referrals (@disproport) and containing
all referrals in the 2018-19 school year for possession of substances.  This
dataset included 30,773 students in grades 6-12 across 45 states. Schools in the
2018-19 dataset numbered 1,903 within 1,008 districts. ODRs are completed within
schools through a standard form generated by a school staff member when students
engage in violations of behavioral rules within the school. Within this form,
fields are completed to indicate circumstances in which the behavior took place
(e.g. date, time, location), characteristics of the undesired behavior (e.g.
behavior category, perceived motivation), student characteristics (e.g.
race/ethnicity, gender), and administrative actions taken (e.g. conference,
suspension, expulsion). In addition, within the dataset, National Center for
Educational Statistics (NCES) data characterizing school population (e.g. school
demographics, population size) and setting (e.g. urbanicity) are paired with
each referral. Student race/ethnicity and disability status data are entered by
the school from enrollment records. The sample contained 66% male students and
34% female students, with 15% of all students in the dataset having IEPs. The
majority of students in the dataset were identified as White (52%), followed by
Latinx students (23%), Black/African American students (10%), students of more
than one race (2%), Indigenous students (2%), Pacific Islander students (1%),
and Asian students (1%). The distribution of students by grade in the dataset
was as follows: grade 6 (7%), grade 7 (14%), grade 8 (20%), grade 9 (21%), grade
10 (18%), grade 11 (12%), and grade 12 (8%). The mean number of referrals per
student was 1.21 (SD = 0.569), with a range of 1-14 referrals for students in
the dataset.

**Table 1**
```{r table_plot}
# cross tabs of race ethnicity by primary substance type
tablesubxrace <- year7 %>%
  group_by(RaceEthnicity, subtype) %>%
  summarise(n = n()) %>%
  spread(subtype, n)
view(tablesubxrace)

kable(
  tablesubxrace,
  format = "html",
  col.names = c("RaceEthnicity", "Alcohol", "Drugs", "Tobacco"),
  align = c("l", "c", "c", "c"),
  caption = "Number of Office Refferals by Race and Substance Type"
  ) %>%
  kable_styling(full_width = TRUE) %>%
  row_spec(row = 0, align = "c") %>%
   footnote(
    general_title = "Note.",
    general = "1 = Native, 2 = Asian, 3 = Latinx, 4 = Black, 5 = White,
    8 = Pacific_Islander, 44 = Multi.",
   footnote_as_chunk = TRUE
    )
same_plot <- ggplot(year7, aes(ODRCount,RaceEthnicity)) +
geom_line() +
facet_wrap(~subtype)
same_plot

rain <- ggplot(year7, aes(x = RaceEthnicity, y = OSS_1stRef)) +
  ggdist::stat_halfeye(
    adjust = .5,
    width = .6, 
    .width = 0, 
    justification = -.2,
    point_colour = NA
  ) + 
  geom_boxplot(
    width = .15,
    outlier.shape = NA
  ) +
  ## add justified jitter from the {gghalves} package
  gghalves::geom_half_point(
    ## draw jitter on the left
    side = "l", 
    ## control range of jitter
    range_scale = .4,
    ## add some transparency
    alpha = .3
  ) +
  coord_cartesian(xlim = c(1.2, NA), clip = "off")
rain
```

### Data Processing:
The following steps were taken to prepare the data set for analysis.
A.	Restricting only the 2018-19 school year.
B.	Creating variables within the referral-level data set.
a.	dummy code use of suspension and expulsion for each referral.
b.	dummy code type of substance possessed for each referral.
C.	Ordering referrals by student ID and date.
D.	Pivoting the data set wider by student so that referrals, in order of receipt,
were nested with students.
E.	Restricting the sample to 6-12 grades.
F.	Restricting variables to those of interest (see variable definition above)
## Missing Data
Missing data was handled by listwise deletion for instances where an individual
student's race or ethnicity was not available (i.e., listed as Not Applicable,
NA). The race or ethnicity identifier was drawn from the National Center on
Educational Statistics approach for collecting and disseminating extant data. In
this case, 7.7% (2360 of 30773 observations) of the data was removed from
analyses because there was insufficient demographic information.
# Measures
## Dependent Variable
  Administrative decisions are actions taken in response to the behavior
violation. Up to five administrative decisions can be listed for each ODR, and
there are 14 options built into the SWIS referral form for administrators to
choose from to describe their decision(s). Options include time in office, loss
of privilege, conference with student, parent contact, timeout/detention,
individualized instruction, in-school suspension (ISS), out of school suspension
(OSS), expulsion, other administrative decision, unknown administrative
decision, bus suspension, and restitution. The definitions for each action taken
are provided in the SWIS training materials provided to stewards of data systems
within each school. To understand the extent to which exclusionary discipline is
used with students possessing substances, referral actions were coded to
indicate whether ISS, OSS, any suspension (coded yes if student received either
ISS or OSS), or expulsion were used with each individual referral. Per the SWIS
training materials, OSS is defined as a 1-3 day period when student is not
allowed on campus (although greater than 3 days was frequently reported under
this code).

## Independent Variables
### Substance Type
  Three substance types are delineated within the SWIS system: Alcohol Possession, Drug Possession, or Tobacco Possession. No additional specification is provided within these codes. For analysis, these codes were turned into dummy-coded variables.
### Race/Ethnicity
  The Race/Ethnicity codes available within this dataset follow federal codes, including codes for Asian, Black/African American, Latinx, Multiracial (more than one race), Indigenous, Pacific Islander, and White. This variable was included as a categorical variable. 

## Independent Variables
### Substance Type
  Three substance types are delineated within the SWIS system: Alcohol Possession, Drug Possession, or Tobacco Possession. No additional specification is provided within these codes. For analysis, these codes were turned into dummy-coded variables.
### Race/Ethnicity
  The Race/Ethnicity codes available within this dataset follow federal codes, including codes for Asian, Black/African American, Latinx, Multiracial (more than one race), Indigenous, Pacific Islander, and White. This variable was included as a categorical variable.

# Research Questions:
1. To what extent does the substance a student possessed for their referral (Alcohol, Tobacco, or Drugs) predict whether or not they receive an out of school suspension for substance possession?
2. To what extent does student race predict whether or not a student receives an out of school suspension for substance possession?
3. To what extent is there an interaction between student race and substance in predicting out of school suspension for substance possession?

# Writeup:
  Descriptive analysis indicates some differences in likelihood of receiving OSS
based on substance type and student race/ethnicity; however, the crosstab table
shows that there are different possession trends by student race/ethnicity. For
all research questions, modeling accounted for variance at the individual
student and school levels. Modeling with nesting at the district level was
considered as well, but over 64% of districts had only one school in the data
set and 85% of districts had two or fewer schools, meaning interpretation of
this variance at the district-level would have limited usefulness if analyzed
with a three-level model.

```{r unconditional_model}
#unconditional models
uncond <- glmer(OSS_1stRef ~ 1 + (1|School), data = year7,
                family = binomial)

# uncond1 <- glmer(OSS_1stRef ~ 1 + (1|School)+ (1|District), data = year7, 
#                 family = binomial)

#look at distribution schools per district: 84% have 2 or fewer
#robust check, remove schools with only 1, run ICC again

summary(uncond)

# translate to probability:
brms::inv_logit_scaled(fixef(uncond))

#69% chance of OSS

# Probability for an individual 1 SD below
brms::inv_logit_scaled(-0.40 - 0.95)
# Probability for an individual 1 SD above
brms::inv_logit_scaled(-0.40 + 0.95)

tidy_uncond <- tidy(uncond, "ran_vals", conf.int = TRUE) %>%
  mutate(level = fct_reorder(level, estimate))

ggplot(tidy_uncond, aes(estimate, level)) +
  geom_linerange(aes(xmin = conf.low, xmax = conf.high),
                 alpha = 0.01) +
  geom_point(color = "#1DA1F2") +
  # get rid of some plot elements
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())

uncond_brm <- brm(OSS_1stRef ~ 1 + (1|School), data = year7,
            family = bernoulli(link = "logit"),
            cores = 4,
            backend = "cmdstanr")

summary(uncond_brm)

pp_check(uncond_brm, type = "bars")

#plot(uncond_bmr)

m0_posterior <- get_parameters(uncond_brm)
head(uncond_brm)

ggplot(m0_posterior, aes(b_Intercept)) +
  geom_density(fill = "#1DA1F2") +
  geom_vline(aes(xintercept = mean(b_Intercept)),
             color = "magenta",
             size = 1.2)
```

# Results:
Suggested length - 1.5 to 2 pages If appropriate, discuss the results of your
model building process and your model comparisons. Include and discuss at least
one table of your model results. Make sure you link specific coefficients or
model results to your research question. Include at least one plot from your
model that communicates your findings (e.g., perhaps comparing the model
predictions to the raw data, or the model predictions for two or more groups)

# Research Question One:
  Analyze use of OSS by substance, logistic multi-level modeling was used
examine differential odds ratios while accounting for variance in decisions
attributable to nesting at the school-level. Substance type variables were
modeled with fixed effects, while variance at the school-level was allowed to
vary as a random effect.
``` {r q1}
#conditional models -substance
# When run all together, tobacco does not sig. predict OSS
drug <- glmer(OSS_1stRef ~  1 + Drugs.1  + (1|School),
             data= year7, 
             family = binomial,
             control = glmerControl(optimizer = "bobyqa")
)
# extract_eq(drug)

alc <- glmer(OSS_1stRef ~  1 + Alcohol.1 + (1|School),
             data= year7, 
             family = binomial)
# extract_eq(alc)

tob <- glmer(OSS_1stRef ~  1 + Tobacco.1 + (1|School),
             data= year7, 
             family = binomial)
# extract_eq(tob)
sub <- glmer(OSS_1stRef ~ 1 + Alcohol.1 + Drugs.1 + Tobacco.1 + (1|School),
      data= year7,
      family = binomial,
             control = glmerControl(optimizer = "bobyqa")
)
# extract_eq(sub)
```

```{r compare_models}
compare_performance(uncond, tob, alc, drug, sub)

# conditional_effects(sub)
summary(sub)
# confint(sub)
# confint keeps throwing an error and I'm not sure why
# exp(coef(sub))

#could compute a mean_diff, like DA did in class. Though, I got a bit lost there too, because we need to know population parameters to input to a bayesian model before building out a likelihood ratio. So I'll start with pseudo code, then show how I might approach it with R

# pred_white <- population_level + year7$RaceEthnicity = White
# pred_latinx <- population_level + year7$RaceEthnicity = latinx
# etc
# sub_medians <- year7 %>%
#   group_by(RaceEthnicity) # %>% 
  # summarize(raceth_medians = median(pred_white, pred_latinx))

# 
# summary(subnonest)
# summary(sub)
# icc(sub)
# confint(sub)
# 
# exp(fixef(drug)) 
# exp(fixef(alc))
# exp(fixef(tob))
# exp(fixef(sub))
# summary(subnonest) #find code for 

# 
# sub1 <- glmer(OSS_1stRef ~  1 + Alcohol.1 + Drugs.1 + (1|School), 
#              data= year7, 
#              family = binomial)
```

```{r plots}
# You are required to create one plot from your model that communicates the coefficients, model predictions, or other features. You are welcome to include more than one, but at least one plot should be created.
```

# Discussion:
  Sample data is from schools implementing PBIS. Average in schools that do not
implement PBIS may be different. It may be appropriate to include other
parameters in models. Only includes out-of-school suspension, and does not
include other forms of exclusion. Nesting at the district-level would provide
answers to questions about how factors at that level (district culture and
policies) influence use of suspension, but with the present data, nesting at
this level would not have provided those answers. The absence of all substances
in a referral (i.e., the intercept) is impossible because the dataset contains
only referrals for substance possession.