---
title: "SeanDavidBrock_HLM2_Final"
author: "Sean Austin, David Fainstein, Brock Rowley"
date: "5/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(here)
library(rio)
library(equatiomatic)
library(easystats)
library(performance)
library(readr)
year7 <- rio::import(here::here("data", "year7_Student1.sav"))
 view(year7)
```
# Data Source: 
National SWIS Referral Data from 2018-19, restricted to only referrals for substance possession, grades 6-12. 

# Research Questions: 
1. To what extent does the substance a student possessed for their referral (Alcohol, Tobacco, or Drugs) predict whether or not they receive an out of school suspension for substance possession?
2. To what extent does student race predict whether or not a student receives an out of school suspension for substance possession?
3. To what extent is there an interaction between student race and substance in predicting out of school suspension for substance possession?

#Data Processing:
This dataset was cleaned up through SPSS. The following steps were taken to prepare the dataset for analysis.
A.	Restricting only the 2018-19 school year
B.	Creating variables within the referral-level dataset to 
a.	dummy code use of suspension and expulsion for each referral
b.	dummy code type of substance possessed for each referral
C.	Ordering referrals by student ID and date
D.	Pivoting the dataset wider by student so that referrals, in order of receipt, were nested with students
E.	Restricting the sample to 6-12 grades
F.	Restricting variables to those of interest (student race/ethnicity, substance type, school ID, district ID, student ID, days of suspension, in school suspension, out of school suspension, any suspension, expulsion)

#Current Challenges:
Create at least 4 models to answer core questions
Understanding first the variance accounted for by school and district nesting
Running models for substance type, race, and then race and substance type combined.

Interpretation of coefficients for substance type will be interesting as well. Currently struggling with interpretation of the intercepts and betas in logistic multi-level modeling. The absence of all substances in a referral (i.e., the intercept) is impossible because the dataset contains only referrals for substance possession.
```{r}
#I think data types (unknown) don't work for MLM, need factors. 
#Tried to turn into factors, but didn't work
year7$PrimarySubstanceType.1 <- as.factor(year7$PrimarySubstanceType.1)

year7 <- year7 %>% 
   mutate(subtype = ifelse(PrimarySubstanceType.1 == 13, "Tobacco", 
                     ifelse(PrimarySubstanceType.1 == 14, "Alcohol", "Drugs"))) %>% 
year7$Alcohol.1 <- as.factor(year7$Alcohol.1) %>% 
year7$Drugs.1 <- as.factor(year7$Drugs.1) %>% 
year7$Tobacco.1 <- as.factor(year7$Tobacco.1) %>% 
year7$RaceEthnicity <- as.factor(year7$RaceEthnicity) %>% 
year7$School <- as.factor(year7$School)


# Getting errors trying to conslidate categories,need to figure out how to make missing
# year7 <- year7 %>% 
#   mutate(RaceEthnicity = ifelse(RaceEthnicity == 1, "Native", 
#                                 ifelse(RaceEthnicity == 2, "Asian", 
#                                 ifelse(RaceEthnicity == 3, "Latinx")))) 
#                                 
#                   ifelse(RaceEthnicity == 4, "Black", 
#                                 ifelse(RaceEthnicity == 5, "White", 
#                                 ifelse(RaceEthnicity == 8, "Pacific_Islander", 
#                                 ifelse(RaceEthnicity == 44, "Multi"))))
#                                 
# 
#                                 # ifelse(RaceEthnicity == 6, is.na(), 
#                                 # ifelse(RaceEthnicity == 7, is.na(), )

#giving me an error for district
year7 <- year7 %>%
  year7$District <- as.factor(year7$District)

#crosstabs race by ethnicity
tablesubxrace <- year7 %>% 
  group_by(RaceEthnicity1, subtype) %>% 
  summarise(n=n()) %>% 
  spread(subtype, n)
view(tablesubxrace)


#unconditional models

uncond <- glmer(OSS_1stRef ~ 1 + (1|School), data = year7, 
                family = binomial)

uncond1 <- glmer(OSS_1stRef ~ 1 + (1|School)+ (1|District), data = year7, 
                family = binomial)

#look at distribution schools per district: 84% have 2 or fewer
#robust check, remove schools with only 1, run ICC again

#conditional models
drug <- glmer(OSS_1stRef ~  1 + Drugs.1  + (1|School)+ (1|District), 
             data= year7, 
             family = binomial)
alc <- glmer(OSS_1stRef ~  1 + Alcohol.1 + (1|School)+ (1|District), 
             data= year7, 
             family = binomial)
tob <- glmer(OSS_1stRef ~  1 + Tobacco.1 + (1|School)+ (1|District), 
             data= year7, 
             family = binomial)
sub <- glmer(OSS_1stRef ~  1 + Alcohol.1 + Drugs.1 + Tobacco.1 + (1|School), 
      data= year7, 
      family = binomial)
subnonest <- glm(OSS_1stRef ~  1 + Alcohol.1 + Drugs.1 + Tobacco.1 , 
                   data= year7, 
                   family = binomial)

summary(subnonest)
summary(sub)
icc(sub)
confint(sub)

exp(fixef(drug)) 
exp(fixef(alc))
exp(fixef(tob))
exp(fixef(sub))
summary(subnonest) #find code for 

# 
# sub1 <- glmer(OSS_1stRef ~  1 + Alcohol.1 + Drugs.1 + (1|School), 
#              data= year7, 
#              family = binomial)
race <- glmer(OSS_1stRef ~  RaceEthnicity + (1|School),
              data= year7, 
              family = binomial)
subxrace <- glmer(OSS_1stRef ~  1 + Alcohol.1 + Drugs.1 + Tobacco.1 + 
                    RaceEthnicity + RaceEthnicity:Alcohol.1 + 
                    RaceEthnicity:Drugs.1 + RaceEthnicity:Tobacco.1 + 
                    (1|School), 
             data= year7, 
             family = binomial) #specify reference group

summary(uncond)
summary(sub)
summary(sub1)
summary(race)
summary(subxrace)

#save race as factor
#vary randomly across school/district as another step
  
summary(uncond1)
icc(sub, by_group = TRUE)
summary(sub1)
?icc

#New dataset restricting to only most diverse schools. 
# Probably only necessary for school-level analysis
racedata <- year7[ which(year7$BLW_Sample==TRUE), ]
view(racedata)
```

